---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#req: 6000 words max. inc ref's 3 figures
#deadline: 1st April
```

##Abstract  

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased. This limits robust inference about any potential “pollination crisis”. [something about how to address the problem]. Analysis of historical time series data is the most direct approach for assessing species declines. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). However, data of this kind are typically rare or non-existent for most pollinator taxa. Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field. We identify opportunities and limitations to progress in assessing pollinator declines at a global scale. Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. Finally, we suggest that there is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally.

## Introduction  

Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction (@ollerton2012) and food security (@kleinetal2007). As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category (@Nieto2014). For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards (@Winfreeetal2011). Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are also extremely diverse, with more than 20,000 species currently described worldwide (@PickeringAscher), and require often require expert taxonomists for identification. The uneven distribution of bee researchers has resulted in geographical biases in bee decline research (@Martinetal2012) and there are also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models [using what parameters?] (@DePalma). A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa (@ref). However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data. The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification (@primak). Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases [including...] (@Keil). But in many cases, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness. In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few “winners” and many “losers” (@Bartomeusetal2013). Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

## The use of museum specimens    
    
Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of collection events (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names (@refs geonames, geoparser and opencage). [Explain zooniverse, NZ initiative and how data can be validated... Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

[how much do we want to extend on this? Is interesting info, but boring o read?]()  
 

##Current evidence on pollinatior declines    

Start with buterflies?

Is not surprising the best evidence of pollinator declines comes from the more disturbed countries western countries. For the UK and the Netherlands, which have been subject to extensive environmental changes, a study using citizen science and museum specimens found declines for bees, flies and flowering plants (@Beisjmieretal2006). However, it is important to note that even for these two countries, the local estimates for pollinator richness are nearby all for large cities and agricultural areas, and data is lacking for conserved natural areas. Further exploration of this dataset showed that the declining trend has attenuated in recent decades (@carvalheiroetal). Using museum specimens in the Netherlands, detailed trends for common bee species confirm the link between pollinator and plant declines, as bees which preferred host plants have declined the most are the more threatened (@Schepper2014).??confusing.    
[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()  

In comparison, regional estimates in species richness changes for the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species tend to persist (@Colla2012?). These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bees from a large forested ecosystem with the few remnants. However, it is important to note that this comparison does not take into account opportunistic bee species using novel land use types in the region. Thus, local richness may have declines but regional richness may not have declined as drastically. In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.  

Outside of Europe and the US there is surprisingly studies on pollinator declines using historical records. [South Africa: Pauw, maybe with plants...]() .  However, there are specific examples showing the historical decline of certain species e.g. Bombus dalbhomi (Aizen)  

[add % in text also]()  

Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.


```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_richness.csv")
#Jamie, if you want, we could complete this data set based on the papers.
# If two estimates available for the same country, I will use two rows. 
# We can compute the mean or the max using reshape here.
# For studies reporting large regions within country this is ok, I think, but
# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
# country, region, individual sites.
#Bee richness can be gathered from Discover life webpage
#head(map)
map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "black",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)

#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))


```
Fig 1: A global map showing evidence (% decline in richness) and Data deficient areas would be great.  


## Beyond species occurrences  
 
Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change (@Bartomeusetal2011). In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions (@tylianakis2010ELE). Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events (@kleinj_raemakers). Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.   
[move this section below]()

## The way forward: Prioritizing the low hanging fruit.   

As illustrated in Figure 1, there is a paucity of countries for which museum data is available, and hence can be used as baseline for assessing pollinator population declines. While ideally, one would aim to digitize all museum records in the world, in the practical world this is an unlikely to happen due mainly to funding constrains. Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.   

Gbif (@ref) is a central repository centralizing occurrence data. Much of this data comes from museum, but it integrates several other sources. Alongside the popular statistical language R (@ref), gbif can be directly queried into your computer (@rgbif) and check data availability for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, organized re-surveys should be organized (e.g. @cameron2013). Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X... Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.   

```{r low hanging fruits, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(rgbif)
apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey

keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
          megachilidae_key, stenotritidae_key, melittidae_key)
#loop through all families, through all countries and store in datarame
country_codes <- isocodes$code 
lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
lhf <- as.data.frame(lhf)
colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
for(i in 1:length(country_codes)){
    for(j in 1:length(keys)){
        lhf[i,1] <- country_codes[i]
        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1980, to = 2018,
                            georeferenced=TRUE,
                            country = country_codes[i]))
        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1850, to = 1980,
                            georeferenced=TRUE,
                            country = country_codes[i]))
    }
}
#very slow, we should save the data and load here when needed.
lhf$post1980 <- rowSums(lhf[,2:8])
lhf$pre1980 <- rowSums(lhf[,9:15])
write.csv(lhf, "data/rgbif_output.csv")
```

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf[order(lhf$pre1980),c(1,17,18)]
plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

```

Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif. Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries hard to get data.    

Alternative plots:  
```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)

plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
#Add country labels in x axes.

plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
abline(h = 8, v = 8)

```
% makes it unidimensional. Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc... Correcting for country size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then?    

Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in occurrence data. We illustrate this process with two contrasted examples for Spain and New Zealand. In the first one, most historical data comes from a few locations and is collected by the same team of people. In this case, we contacted the original authors to understand their sampling protocol, and resurveyed the same sites 20 years later. The second case shows a diversity of collectors and sites. In this case, we should ensure that geographical and taxonomic coverage do not change systematically with time. For example, some collectors may target only certain taxa, biasing the sample. We provide annotated R scripts to treat both kind of data while revealing trends for two relevant regions with contrasting histories. We hope this resource will encourage people to add more data and analysis... (include a repository where people can contribute?)   

## Case study one: Spain  

Spain is an interesting study system because is has been transformed by humans heavely, but is not as intensified as other EU countries. In addition, Spain is a hotspot of bee diversity and maintains an heterogeneous landscape. As illustrated in Figure 2, Spain has already digitalized a decent amount of occurrence data both historical, and recent. However, a visual inspection reveals that data points are clustered and historical and recent records do not match spatially, making them hard to compare. Among the old records, it is striking that most of them are located around Valladolid and collected by Enrique Asensio and his team. Unfortunately, there is no recent sampling in the area. After some investigations, we discover that Enrique sampled systematically 7 localities and that more data is available at the "Museo de historia Nacional" and other minor collections. An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 7 sites.   
 
In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test. [Maybe more analysis arise...](). We show that ...    

Fig 3: Mutipanel figure  

##Case study two: NZ    
New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation. Much of New Zealand’s pollinator fauna is also relatively depauperate. For example, New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c. 1600 species (REF). However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005). Therefore, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.

Human colonisation of New Zealand was relatively recent in global terms (c. 740 y) (@Wilmshurst et al. 2008). Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014). Therefore, human activity likely affected New Zealand pollinator communities long before surveys and collections of pollinators began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

New Zealand pollinator collections
We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections. Fly pollinator data was obtained from six participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators. Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).

Data analyses
First, we filtered our original datasets so that data for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through time by binning collection records, so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details). We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. (should we analyse natives and exotics together or separately?). For species that showed overdispersion, we used a quasi-binomial distribution. We only included species in this analysis that had more than 30 records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year. 

Changes in species richness over time 
We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05). In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).

Changes in species relative abundances over time 
We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic). Three bee species declined in relative abundance (one native and two exotic). However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).

Fig 3: Multi-panel figure

-	Do we need some sort of separate discussion or integrate into the results section?
… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.


## The way forward  

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. We show in our case studies that collaboration with both museum curators and data collectors is key to understanding the data and treating it appropriately. We envision that rather than a central repository of museum collection records, we should aim to make digitalized data readily available and easy to share. Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access. However, even when data is stored in different webpages or databases, it can be retried and joined with other data sets using open science tools (@ropensci) if it is easily accessible and has programmatic access (API e.g. @heltmint).   

We also need to revolutionize the way that researchers collaborate with museums by foresting healthy bidirectional relationships. For instance, researchers collect massive numbers of specimens, but these are often inappropriately vouched, rendering them useless for research purposes. To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding and recognition that accurate data recording and long term preservation are critical for research.
[What about an interactive bee decline map? or Integration of DATA and analysis:]()  

Ultimately, to identify global trends in pollinator declines we require data from diverse geographic regions. In particular, we need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe. Furthermore, we need to investigate whether there have been declines in pollinator groups other than bees. As far as we know, there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems of the US and Europe (@Bartomeus, @dePalma, @Carvalheiro). For most other pollinator taxa we know almost nothing. For many taxa, the first step is to identify and describe them and only then can we begin to document declines.  

#Acknowledgements  

We thank E. Asensio (or maybe coauthor?), Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.  



