---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#req: 6000 words max. inc ref's 3 figures
#deadline: 1st April
```

##Abstract  

There is ever-increasing concern about the decline of pollinators, especially for bees. However, despite reports that pollinator declines are widespread, data is scarce and geographically and taxonomically biased. This limits any robust inference about a potential pollination crisis. The most direct approach for identifying patterns in species declines is analysis of historical time series data. However, these data are often rare or non-existent for most pollinator groups, but provide the only source of historical information serving as a baseline to understand the extent of pollinator declines. Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current occurrence patterns to identify changes in biodiversity, relative species abundances, body size and even pollination services over time. Despite recent advances, we show that information on the status and trends of most pollinators is missing and a key bottle neck slowing progress in this field is the digitization of historical data. We identify opportunities and limitations to progress understanding of pollinator declines worldwide. Further, we demonstrate how to analyse historical museum collection data, by providing two contrasting case studies from distinct geographical regions, New Zealand and Spain, for which historical pollinator declines have never been assessed. We show that there is a great potential for museum specimens to play a key role in identifying the extent of the pollination crisis if focus on unleashing the power of historical collection data.

## Introduction  

Animal pollinators are acritical component of ecosystems, given their role in plant reproduction (@ollerton2012) and food security worldwide (@kleinetal2007). As many other taxa, pollinators are vulnerable to multiple human-mediated environmental changes, which potentially threatening their populations. This problem was identified by researchers several decades ago, and was popularized by the influential book "the forgotten pollinators" (1996). However, these accounts were relatively anecdotal as there were almost no data on pollinator trends and statuses at the time. These initial claims triggered the first efforts to address the problem, including a US National Academy of Science (NAS) panel commissioned to assess the extent of pollinators declines (2006). The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to global change have multiplied and researchers have reached strong consensus that pressures such as habitat destruction, land use change, chemical exposure, exotic species and climate change are the primary causes of pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the status and trends of most pollinator species worldwide remain unknown. For example, even for a well-studied region such as Europe, 60% [check number]() of bee species assessed by in recent IUCN report fall into the category "Data Deficient" (@Nieto2014). For countries outside of Europe and the US, data on pollinator population is almost non-existent. 

One of the main barriers to identifying pollinator population trends is that pollinators are a diverse group of taxa including bees, flies, butterflies and beetles, and also birds, bats and even lizards (@Winfreeetal2011). In particular, bees are usually regarded as the more important pollinaotrs due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are an incredibly diverse group, with more than 20,000 species worldwide (@PickeringAscher), and most require taxonomic expertise for identification. This has led to a biased geographic patterns in bee decline research due to the uneven distribution of researchers (@Martinetal2012), but also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier is methodological and involves challenges for assessing population trends for highly mobile organisms, which are also often short-lived and small.
One solution to overcoming these barriers has been to use space for time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data currently available and in Europe, researchers have recently estimated bee richness declines for each country using predictions generated from models (@DePalma). A second promising method is pollinator monitoring programs, with data often collected by citizen scientists. This approach was inspired by successful butterfly monitoring programs and is currently being extended to other pollinator taxa (@ref). The main limitation is that these programs need significant time to generate long-term datasets and do not reveal anything about historic pollinator communities. Finally, the most robust approach for identifying historical pollinator tends is to use museum collection data. The core aim of museums is precisely to conserve and curate collections, thus they not only serve as a precious repository specimens, but often also ensure high quality taxonomic identification (@primak). However, the major bottleneck for researchers wanting to use this information is the digitization of specimens. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have facilitated identification of current status and trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases (@Keil). In reality, museum collections are the only data we have to reconstruct the past. 

Regardless of the method, it is important to consider that biodiversity has many aspects beyond species richness, and we should into identifying changes not only in local (alpha) diversity or regional (gamma) diversity, but also in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few winners and many of losers (@Bartomeusetal2013). Identifying winners and losers is critical as the few winners are often exotic species, and represent a subset of trait that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

## The use of museum specimens    
    
Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of the collection event (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens a museum collection is a daunting task and requires specialized personnel. While some tasks can only be undertaken by those with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the process of digitizing collections. High resolution photos of specimens and the associated labels can be uploaded to the internet, where the work of transcribing the images can be distributed across hundreds or thousands of volunteers. Furthermore, new algorithms have been created to geo-reference localities based on vernacular names (@refs geonames, geoparser and opencage). Explain zooniverse, NZ initiative and how data can be validated... [Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]
[how much do we want to extend on this? Is interesting info, but boring o read?]()  

##Current evidence on pollinatior declines    

Start with buterflies?

Is not surprising the best evidence of pollinator declines comes from the more disturbed countries western countries. For the UK and the Netherlands, which have been subject to extensive environmental changes, a study using citizen science and museum specimens found declines for bees, flies and flowering plants (@Beisjmieretal2006). However, it is important to note that even for these two countries, the local estimates for pollinator richness are nearby all for large cities and agricultural areas, and data is lacking for conserved natural areas. Further exploration of this dataset showed that the declining trend has attenuated in recent decades (@carvalheiroetal). Using museum specimens in the Netherlands, detailed trends for common bee species confirm the link between pollinator and plant declines, as bees which preferred host plants have declined the most are the more threatened (@Schepper2014).??confusing.    
[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()  

In comparison, regional estimates in species richness changes for the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species tend to persist (@Colla2012?). These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bees from a large forested ecosystem with the few remnants. However, it is important to note that this comparison does not take into account opportunistic bee species using novel land use types in the region. Thus, local richness may have declines but regional richness may not have declined as drastically. In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.  

Outside of Europe and the US there is surprisingly studies on pollinator declines using historical records. [South Africa: Pauw, maybe with plants...]() .  However, there are specific examples showing the historical decline of certain species e.g. Bombus dalbhomi (Aizen)  

[add % in text also]()  

Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.


```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_richness.csv")
#Jamie, if you want, we could complete this data set based on the papers.
# If two estimates available for the same country, I will use two rows. 
# We can compute the mean or the max using reshape here.
# For studies reporting large regions within country this is ok, I think, but
# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
# country, region, individual sites.
#Bee richness can be gathered from Discover life webpage
#head(map)
map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
#have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "black",
               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)

#add a modified legend using the same initial parameters as mapCountryData
do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))


```
Fig 1: A global map showing evidence (% decline in richness) and Data deficient areas would be great.  


## Beyond species occurrences  
 
Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change (@Bartomeusetal2011). In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions (@tylianakis2010ELE). Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events (@kleinj_raemakers). Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.   
[move this section below]()

## The way forward: Prioritizing the low hanging fruit.   

As illustrated in Figure 1, there is a paucity of countries for which museum data is available, and hence can be used as baseline for assessing pollinator population declines. While ideally, one would aim to digitize all museum records in the world, in the practical world this is an unlikely to happen due mainly to funding constrains. Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.   

Gbif (@ref) is a central repository centralizing occurrence data. Much of this data comes from museum, but it integrates several other sources. Alongside the popular statistical language R (@ref), gbif can be directly queried into your computer (@rgbif) and check data availability for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, organized re-surveys should be organized (e.g. @cameron2013). Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X... Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.   

```{r low hanging fruits, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(rgbif)
apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey

keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
          megachilidae_key, stenotritidae_key, melittidae_key)
#loop through all families, through all countries and store in datarame
country_codes <- isocodes$code 
lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
lhf <- as.data.frame(lhf)
colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
for(i in 1:length(country_codes)){
    for(j in 1:length(keys)){
        lhf[i,1] <- country_codes[i]
        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1980, to = 2018,
                            georeferenced=TRUE,
                            country = country_codes[i]))
        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1850, to = 1980,
                            georeferenced=TRUE,
                            country = country_codes[i]))
    }
}
#very slow, we should save the data and load here when needed.
lhf$post1980 <- rowSums(lhf[,2:8])
lhf$pre1980 <- rowSums(lhf[,9:15])
write.csv(lhf, "data/rgbif_output.csv")
```

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf[order(lhf$pre1980),c(1,17,18)]
plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

```

Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif. Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites. Lower right are sites which needs to dig museum records and lower left are countries hard to get data.    

Alternative plots:  
```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)

plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
#Add country labels in x axes.

plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
abline(h = 8, v = 8)

```
% makes it unidimensional. Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc... Correcting for country size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then?    

Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in occurrence data. We illustrate this process with two contrasted examples for Spain and New Zealand. In the first one, most historical data comes from a few locations and is collected by the same team of people. In this case, we contacted the original authors to understand their sampling protocol, and resurveyed the same sites 20 years later. The second case shows a diversity of collectors and sites. In this case, we should ensure that geographical and taxonomic coverage do not change systematically with time. For example, some collectors may target only certain taxa, biasing the sample. We provide annotated R scripts to treat both kind of data while revealing trends for two relevant regions with contrasting histories. We hope this resource will encourage people to add more data and analysis... (include a repository where people can contribute?)   

## Case study one: Spain  

Spain is an interesting study system because is has been transformed by humans heavely, but is not as intensified as other EU countries. In addition, Spain is a hotspot of bee diversity and maintains an heterogeneous landscape. As illustrated in Figure 2, Spain has already digitalized a decent amount of occurrence data both historical, and recent. However, a visual inspection reveals that data points are clustered and historical and recent records do not match spatially, making them hard to compare. Among the old records, it is striking that most of them are located around Valladolid and collected by Enrique Asensio and his team. Unfortunately, there is no recent sampling in the area. After some investigations, we discover that Enrique sampled systematically 7 localities and that more data is available at the "Museo de historia Nacional" and other minor collections. An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 7 sites.   
 
In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test. [Maybe more analysis arise...](). We show that ...    

Fig 3: Mutipanel figure  

##Case study two: NZ    
New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation. Much of New Zealand’s pollinator fauna is also relatively depauperate. For example, New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c. 1600 species (REF). However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005). Therefore, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.

Human colonisation of New Zealand was relatively recent in global terms (c. 740 y) (@Wilmshurst et al. 2008). Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014). Therefore, human activity likely affected New Zealand pollinator communities long before surveys and collections of pollinators began. Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.

New Zealand pollinator collections
We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections. Fly pollinator data was obtained from six participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators. Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).

Data analyses
First, we filtered our original datasets so that data for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through time by binning collection records, so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details). We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link. (should we analyse natives and exotics together or separately?). For species that showed overdispersion, we used a quasi-binomial distribution. We only included species in this analysis that had more than 30 records. To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year. 

Changes in species richness over time 
We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05). In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).

Changes in species relative abundances over time 
We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic). Three bee species declined in relative abundance (one native and two exotic). However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).

Fig 3: Multi-panel figure

-	Do we need some sort of separate discussion or integrate into the results section?
… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.


## The way forward  

As illustrated in this examples, unleashing the power of museum data to answer pressing questions is at our hands, and need the coordinated effort of many actors. As evidenceed by the case studies, collaborating with museum curators and original data collectors is key to understand the data and treat it appropiatelly. We envision that rather than a central repository of museum records, we should aim to make the digitalized data available and easy to share, and if possible ultimately centralized in existing platforms like Gbif. Even when the data lives in different museum webpages or databases, if it's easily accessible and has programmatic access (API e.g. @heltmint) it can be retried and joined with other data sets using open science tools (@ropensci).   

We also need to revalorize museums and foster a birectional relationship between researchers and museums. Researchers collect massive amounts of specimens, not always appropriately vouched. Depositing this in museums should ensure its long term preservation. However, this require appropriately funded museums and recognizing data (and not only papers) as an important research output.  

[What about an interactive bee decline map? or Integration of DATA and analysis:]()  

Ultimately, we need more data to compare global differences geographically and see which are the generalized trends. As far as we know there have been few bee regional extinctions (@Ollerton, @Bartomeus2013), but bee declines are widespread in disturbed ecosystems of north america and Europe (@Bartomeus, @dePalma, @Carvalheiro). We know almost nothing about the rest... for those areas, the first step is to know what pollinators exists and name the species before we loss them.  

#Acknowledgements  

E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL. Barry Donovan for providing New Zealand bee collection records. Carola Warner for entering the New Zealand data.  


