---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#req: 6000 words max. inc ref's 3 figures
#deadline: 1st April
```

##Abstract

Statements about pollinator crisis, or bee declines are widespread, but information is currenti limiting....
Building on previous studies showing the use of historical museum specimens to compare historical bee occurrences in northeast USA, Central Europe, and South Africa with current occurrence patterns, we show the potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations. Highlight knowledge gaps and oportunities (e.g. London museum, zooniverse...). Interestingly, we will add data from New Zealand and Spain increasing the geographical coverage and including bee diversity hotspots like Spain, for which bee declines have never been assessed.


## Intro

Paragraph on pollinator importance and claims of a crisis. initial claims driving to NAS 2007.  Goulson Science, Gonzalez -varo TREE, Potts TREE. BUT highlight not global assessment and doubts (Gazoul "buzzines as usual", Forrest or Rafferty Oikos paper on geograpfical bias). Taxonomic bias?

Define trends: Decline, turnover (changes in evennes winners loosers), exotics... Review current methods to assess 1) space for time (recent reviews rae, ...), 2) monitoring schemes /citicen science (iNaturalis) (UK) and 3) museums type data. Highlight its importance (cite Primak). Pros and cons with an enphasis on Museums. 

##Current evidence:
- US: Bartomeus 2013 and Burkle 201?, ... NAS book Critizize Burkle dessign and compare to mine.
- EU: Beijsmier, Carvalho, Schepper, ... UK? Peter Keil flies? Moths? Buterflies!
- South Africa: Pauw, maybe with plants...
- Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters

We compiled current evidence worldwide and show that ... (short methods included in text)

```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(rworldmap)
map <- read.csv("data/bee_declines.csv", sep = ";")
#Jamie, if you want, we could complete this data set based on the papers.
# If two estimates available for the same country, I will use two rows. 
# We can compute the mean or the max using reshape here.
# For studies reporting large regions within country this is ok, I think, but
# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
# country, region, individual sites.
#Bee richness can be gathered from Discover life webpage
#head(map)
map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
                            nameCountryColumn = "distribution")
#need to use ISO's to make sure we don't carry over NA's due to country misspellings
mapCountryData(mapToPlot = map2, nameColumnToPlot = "bee_richness"
               , mapTitle = "Bee richness", addLegend = FALSE) 

```

Fig 1: a global map showing evidence (% decline in richness) and Data deficient areas would be great.


## Beyond species occurrences
Other uses of museums in pollinators. Identifying mechanisms of decline
- Oliveira plos one 
- Osorio Ecol Lettres
- Phenoloy (Bartomeus)


## Prioritizing the low hanging fruit.

Analyze trade off between number of records in Gbif for > 1980 and pre <1980. Each quadrant needs an analysis. 
+ we provide code to run it, and maybe add it to the repo per country? distributed database style?
We do it for two regions:


```{r low hanging fruits, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(rgbif)
apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey

keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
          megachilidae_key, stenotritidae_key, melittidae_key)
#loop through all families, through all countries and store in datarame
country_codes <- isocodes$code 
lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
lhf <- as.data.frame(lhf)
colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
for(i in 1:length(country_codes)){
    for(j in 1:length(keys)){
        lhf[i,1] <- country_codes[i]
        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1980, to = 2018,
                            georeferenced=TRUE,
                            country = country_codes[i]))
        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
                            from = 1850, to = 1980,
                            georeferenced=TRUE,
                            country = country_codes[i]))
    }
}
#very slow, we should save the data and load here when needed.
lhf$post1980 <- rowSums(lhf[,2:8])
lhf$pre1980 <- rowSums(lhf[,9:15])
write.csv(lhf, "data/rgbif_output.csv")
```

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf[order(lhf$pre1980),c(1,17,18)]
plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

```

Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif. Upper rigth cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites. Lower right are sites which needs to digg museum records and lower left are countries hard to get data.

I am not sure the 1980 cut is apropiate. Looks like it is too correlated. Maybe we should correct per country size? Or per total records? % post vs % pre.

```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)

plot(lhf$percent_post1980, lhf$percent_pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,1), ylim = c(0, 1))
abline(h = 0.50, v = 0.50)

plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
abline(h = 8, v = 8)

```
This is even worst... correcting for size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then??
Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc...


We ilustrate how this can be done using two countries and providing code to reproduce our analysis, so that more countries can contribute.

## Case study one: Spain

- Describe data and analysis.
- Present main results

## Case study two: NZ

- Describe data and analysis.
- Present main results


## Future needs /knowledge gaps:
- Digitalization of MORE museums (coordinated) Just common protocols? RopenSci? What about an interactive bee decline map? Predicts?
- Integration of DATA and analysis:
- Communication Museums- Sci (biderictional. Importance of voucering specimens fused for science).
- Compare global differences geographically (Eu Z US, for example)


## Conclusions: Joint efforts needed at the global scale, scit science, ...

highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably (central africa).


#Acknowledgements

E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL.
